### Load required packages

```{r}
#load packages   
library(terra)  
library(tidyverse) 
library(sf) 
library(PROJ) 
```

### Data sources & import:

1.  CoCT BioNet with Habitat condition
2.  National Vegetation Map 2024 version, vector data (ESRI file geodatabase) January 2025 curated by SANBI [@rutherford2006; @Dayaram2019].
3.  National Land Cover 2022 7 class version prepared by SANBI.

```{r}
# Add City of Cape Town Land COver and Condition 2023 (prepared in ARCGIS PRO), where 0 = inside AOI but no value assigned, 4 = notnat, 2 = sec nat, 8 = severely degraded. 
coct <- rast("C:/Users/skownoa/Dropbox/NBAwork/RLEworking/CoCT/coct_cond10.tif")

# Add national LC, resample to match 20m AEA grid used for supplementary analysis
nlc <- rast("C:/Users/skownoa/Dropbox/NBAwork/NLC_change/nlc2022_7class.tif" )
nlc <- resample(nlc, coct, method = "near") #  match extent 

# Add vegetation data  
veg <- vect("C:/GIS/NBA/VegMap/VEGMAP2024/NVM2024final/NVM2024final.gdb", "NVM2024Final_IEM5_12_07012025")  

# Clean up non veg names 
veg$T_MAPCODE <- ifelse(veg$T_Name == "Non-terrestrial (estuarine habitat)" | veg$T_Name == "Non-terrestrial (Aquatic)", "NOTVEG", veg$T_MAPCODE)   

# rasterise the veg vector 
veg_r <- rasterize(veg, nlc, "T_MAPCODE")  
```

## Analysis

*Cross-tabulate land cover time series with vegetation.*

Step 1. Stack the land cover and condition data for COCT, NL and vegetation.

Step 2. Cross-tabulate raster stack to the output table (in tibble format).

```{r}
# Analysis #Step 1-2.   # get LC  at 1990 time period 
#coct_rall  <- as_tibble(crosstab(c(nlc, veg_r, coct)))   
#write.csv(coct_rall, file = "outputs/coct_rall.csv")

## read in data if spatial analysis step is being skipped
coct_rall <- read.csv("C:/Rdata/LCC_terr/outputs/coct_rall.csv")

# clean up names and add km2
coct_rall  <- coct_rall  %>%    
  mutate(km2 = n*400/1000000) %>% # pixel size is 20m
  rename(LC = nlc2022_7class) %>%
  rename(COCT = coct_cond10)  

# Calculate extent of each vegtype within the specific AOI 
coct_list <- coct_rall %>%
  filter(COCT != 99, T_MAPCODE != "NOTVEG") %>%
  group_by(T_MAPCODE) %>%
  summarise(ext_coct = sum(km2))

# summarise
coct_sum <- coct_rall %>%
  filter(T_MAPCODE != "NOTVEG") %>%
  mutate(LCcomb = case_when(
                            COCT == 1 ~ "nat",
                            COCT == 2 | COCT == 4 ~ "notnat",
                            COCT == 8 ~ "deg80",
                            COCT == 99 & LC == 1 ~ "nat",
                            COCT == 99 & LC > 1 ~ "notnat",
                            TRUE ~ "unknown")) %>%
  group_by(T_MAPCODE, LCcomb) %>%
  summarise(km2tot = sum(km2)) %>%
  pivot_wider(names_from = LCcomb, values_from = km2tot) %>%
    mutate(prcdeg_coct = 100*round(deg80/(nat+deg80),2)) %>% # cond of remnants (D1)
    mutate(prclost_coct = 100* round(notnat/(nat+notnat+deg80),2)) %>% # ext of remnants (A3)
    mutate(D3_coct = case_when(prcdeg_coct >= 90 ~ "CR",
                   prcdeg_coct >= 70 ~ "EN",
                   prcdeg_coct >= 50 ~ "VU",
                  TRUE ~ "LC"))  %>%
  mutate(A3_coct = case_when(prclost_coct >= 90 ~ "CR",
                   prclost_coct >= 70 ~ "EN",
                   prclost_coct >= 50 ~ "VU",
                  TRUE ~ "LC")) %>%
      left_join(coct_list, by = "T_MAPCODE") %>%
   mutate(ext_coct = ext_coct/(nat + notnat + deg80))

write.csv(coct_sum, file = "outputs/coct_sum.csv")
```
