---
bibliography: references.bib
---

### Load required packages

```{r}
#load packages   
library(terra)  
library(tidyverse) 
library(sf) 
library(PROJ) 
```

### Data sources & import:

1.  Regional land degradation assessment for the Hardeveld Bioregion conducted by Bell et al in 2021 [@bell2021].

2.  National Land Cover (7 class version prepared by SANBI) 2022.

3.  National Vegetation Map 2024 version, vector data (ESRI file geodatabase) January 2025 curated by SANBI [@rutherford2006; @Dayaram2019].

```{r}
# Add Hardeveld degradation data (prepared in ARCGIS PRO), where 0 = inside AOI but level of degradation < that 80%; where 8 = severely degraded >= 80%). 
HV <- rast("C:/Users/skownoa/Dropbox/NBAwork/RLEworking/Alternative_LC_DEG/hv_bell.tif")

# Add national LC, resample to match 20m AEA grid used for supplementary analysis
nlc <- rast("C:/Users/skownoa/Dropbox/NBAwork/NLC_change/nlc2022_7class.tif" )
nlc <- resample(nlc, HV, method = "near") #  match extent 

# Add vegetation data  
veg <- vect("C:/GIS/NBA/VegMap/VEGMAP2024/NVM2024final/NVM2024final.gdb", "NVM2024Final_IEM5_12_07012025")  

# Clean up non veg names 
veg$T_MAPCODE <- ifelse(veg$T_Name == "Non-terrestrial (estuarine habitat)" | veg$T_Name == "Non-terrestrial (Aquatic)", "NOTVEG", veg$T_MAPCODE)   
# rasterise the veg vector 
veg_r <- rasterize(veg, nlc, "T_MAPCODE")  
```

## Analysis

*Cross-tabulate land cover time series with vegetation.*

Step 1. Stack the land cover for each time period with the vegetation data.

Step 2. Cross-tabulate each raster stack and add a column denoting the YEAR (from Land Cover) to the output table (in tibble format). Rename each output column as "LC" to allow binding later.\

```{r}
# Analysis #Step 1-2.   # get LC  at 1990 time period 
hv_rall  <- as_tibble(crosstab(c(nlc, veg_r, HV)))   
write.csv(hv_rall, file = "outputs/hv_rall.csv")

# Import cross tab output to avoid running spatial analysis 
#hv_rall <- read.csv("C:/Rdata/LCC_terr/outputs/hv_rall.csv")

# Calculate extent of each vegtype within the specific AOI 
hv_list <- hv_rall %>%
  filter(hv_bell != 99, T_MAPCODE != "NOTVEG") %>%
  group_by(T_MAPCODE) %>%
  summarise(ext_hv = sum(n)*400/1000000)

# clean up names and add km2
hv_rall  <- hv_rall  %>%    
  mutate(km2 = n*400/1000000) %>% # pixel size is 20m
  rename(LC = nlc2022_7class) %>%
  rename(HV = hv_bell)

options(scipen =999)
hv_sum <- hv_rall %>%
  filter(T_MAPCODE != "NOTVEG") %>%
  mutate(LCcomb = case_when(
                            HV == 8 ~ "deg80", 
                            LC == 1 ~ "nat", 
                            LC > 1 ~ "notnat", 
                            TRUE ~ "unknown")) %>%
  group_by(T_MAPCODE, LCcomb) %>%
  summarise(km2tot = sum(km2)) %>%
  pivot_wider(names_from = LCcomb, values_from = km2tot) %>%
    mutate(prcdeg_hv = 100*round(deg80/(nat+deg80),2)) %>% # cond of remnants (D3)
    mutate(D3_hv = case_when(prcdeg_hv >= 90 ~ "CR",
                   prcdeg_hv >= 70 ~ "EN",
                   prcdeg_hv >= 50 ~ "VU",
                  TRUE ~ "LC")) %>%
    left_join(hv_list, by = "T_MAPCODE") %>%
   mutate(ext_hv = ext_hv/(nat + notnat + deg80))

write.csv(lk_sum, file = "outputs/hv_sum.csv")
  
```
