---
title: "Land Cover Change - South Africa, Lesotho, Eswatini 1990-2022"
subtitle: "Change in LCLU (7 class) per terrestrial ecosystem type, including Eswatini and Lesotho"
author: "SANBI"
date: January  2025
format: 
 html:
  toc: true
  toc-location: left
  embed-resources: true
editor: visual
execute: 
  eval: false
bibliography: references.bib
---

### Approach

Cross tabulate terrestrial ecosystem type (vegetation) data with land cover data 1990-2022 for South Africa. Two versions of this analysis have been developed. This version is focused on producing metrics for the Red List of Ecosystems assessments, as such it contains ecosystem extent and land cover data for Eswatini and Lesotho. There is an alternative version for which these countries are masked out and national metrics for South Africa are computed (for EAI & LC change information). These data sets are aligned to same extent, resolution and projection. They include land cover for Eswatini and Lesotho based on ESA Africa 2016 landcover product (i.e., no true land ocover change data).

### Load required packages

```{r}
#load packages  
library(terra) 
library(tidyverse)
library(sf)
library(PROJ)

```

### Data sources & import:

1\. South African National Land Cover data sets for 1990, 2014, 2018, 2020 and 2022 (prepared by the National Department of Forestry, Fisheries and the Environment) were adjusted to have a common raster resolution (20m) and snapped to the extent of 2022 data. The data were reclassified in ARCGIS PRO into seven classes: 1 = Natural; 2 = Secondary Natural, 3 = Artificial water bodies, 4 = Built up, 5 = Croplands, 6 = Mines, 7 = Plantation. These data sets are aligned to same extent, resolution and projection. They include land cover for Eswatini and Lesotho based on ESA Africa 2016 land cover product (i.e., no true land cover change data).

2\. National Vegetation Map 2024 version, vector data (ESRI file geodatabase) January 2025 curated by SANBI [@rutherford2006; @Dayaram2019].

```{r}

#Add data 

# the data are available at this link 
# https://www.dropbox.com/scl/fo/tcmdhlggiaq3d2brvize7/AHvcF04czyJvGnyZSwqfK5w?rlkey=ua3ollzupol34ditiqqvsdcw0&dl=0 
#download them to a location on the local machine then adjust the file paths as required  

lc1990 <- rast("C:/Users/skownoa/Dropbox/NBAwork/NLC_change/nlc1990_7class.tif") 
lc2014 <- rast("C:/Users/skownoa/Dropbox/NBAwork/NLC_change/nlc2014_7class.tif")
lc2018 <- rast("C:/Users/skownoa/Dropbox/NBAwork/NLC_change/nlc2018_7class.tif")
lc2020 <- rast("C:/Users/skownoa/Dropbox/NBAwork/NLC_change/nlc2020_7class.tif")
lc2022 <- rast("C:/Users/skownoa/Dropbox/NBAwork/NLC_change/nlc2022_7class.tif")

#simplify to test__________________________________________________________

# agg_fact <- 10
# mode_function <- function(x) {
#     ux <- unique(x)
#     ux[which.max(tabulate(match(x, ux)))]
# }
# 
# lc2022 <- aggregate(lc2022, fact = agg_fact, fun = mode_function)
# lc2020 <- aggregate(lc2020, fact = agg_fact, fun = mode_function)
# lc2020 <- resample(lc2020, lc2022, method ="near")
# lc2018 <- aggregate(lc2018, fact = agg_fact, fun = mode_function)
# lc2018 <- resample(lc2018, lc2022, method ="near")
# lc2014 <- aggregate(lc2014, fact = agg_fact, fun = mode_function)
# lc2014 <- resample(lc2014, lc2022, method ="near")
# lc1990 <- aggregate(lc1990, fact = agg_fact, fun = mode_function)
# lc1990 <- resample(lc1990, lc2022, method ="near")
#____________________________________________________________________________

#load vegetation data 
veg <- vect("C:/GIS/NBA/VegMap/VEGMAP2024/NVM2024final/NVM2024final.gdb", "NVM2024Final_IEM5_12_07012025")

# to help differentiate masked areas (LES and SWA) from NON TERRESTRIAL areas add "NOTVEG" to T_MAPCODE where T_NAME = "Non-Terrestrial*"
veg$T_MAPCODE <- ifelse(veg$T_Name == "Non-terrestrial (estuarine habitat)" | veg$T_Name == "Non-terrestrial (Aquatic)", "NOTVEG", veg$T_MAPCODE) 

# rasterise the veg vector
vegrast <- rasterize(veg,lc2022,"T_MAPCODE") 

```

## Analysis

*Cross-tabulate land cover time series with vegetation.*

Step 1. Stack the land cover for each time period with the vegetation data.

Step 2. Cross-tabulate each raster stack and add a column denoting the YEAR (from Land Cover) to the output table (in tibble format). Rename each output column as "LC" to allow binding later.

***Note this process takes approximately 2 hours per cross tabulation.***

```{r}
# Analysis #Step 1-2.  
# get LC  at 1990 time period
tb1990 <- as_tibble(crosstab(c(lc1990,vegrast)))  
tb1990 <- tb1990 %>% 
  add_column(YEAR = 1990) %>% 
  rename(LC = nlc1990_7class) 
# get LC at 2014 time period
tb2014 <- as_tibble(crosstab(c(lc2014,vegrast)))  
tb2014 <- tb2014 %>% 
  add_column(YEAR =2014) %>% 
  rename(LC = nlc2014_7class) 
# get LC at 2018 time period
tb2018 <- as_tibble(crosstab(c(lc2018,vegrast)))  
tb2018 <- tb2018 %>% 
  add_column(YEAR =2018) %>% 
  rename(LC = nlc2018_7class)  
# get LC  at 2020 time period
tb2020 <- as_tibble(crosstab(c(lc2020,vegrast)))  
tb2020 <- tb2020 %>% 
  add_column(YEAR =2020) %>% 
  rename(LC = nlc2020_7class) 
# get LC  at 2020 time period
tb2022 <- as_tibble(crosstab(c(lc2022,vegrast)))  
tb2022 <- tb2022 %>% 
  add_column(YEAR =2022) %>% 
  rename(LC = nlc2022_7class) 

# build a table for the 1750 LC where all is natural
tb1750 <- tb1990 %>%
 filter(T_MAPCODE != '<Null>'& T_MAPCODE != " " & LC != 0) %>%
 mutate (YEAR = 1750) %>%
 group_by (YEAR, T_MAPCODE) %>%
 summarise(n = sum(n) , .groups = "drop") %>%
 mutate(LC = "1") %>% # this in "" because  other LC are char for some reason
 relocate(YEAR, .after = LC)%>%
 relocate(LC, .before = T_MAPCODE)

```

### Build results table

Combine (bind) the output tables and convert pixel counts to extent in square kilometers. Pixels in these rasters are 20m x 20m (400 x400 in the test), so conversion requires multiplication by cell area (400m2) (or160000 in test) and division by 1000 000 to result in km2. Add BIOME column using T_MAPCODE prefixes. Clean up field names and remove rows with Null T_MAPCODE values (small areas where vegmap vector includes non terrestrial ecosystems, such as lakes, estuaries). Due to earlier steps non terrestrial areas are retained and have NOTVEG as T_MAPCODE and BIOME . Add text labels for each land cover class.

```{r}
# Step 5. add extent based on pixel size and then make raster binary by changing 4 to 2 (not natural pixels) and 3 to 1 (natural pixels) 
lc7_list = list(tb1750,tb1990,tb2014,tb2018,tb2020,tb2022) 
lc7_rall <- lc7_list %>%  bind_rows() %>%   
  as_tibble() %>% 
  filter(T_MAPCODE != '<Null>'& T_MAPCODE != " " & LC != 0) %>%
  group_by(LC, YEAR, T_MAPCODE) %>%
  summarise(n = sum(n), .groups = "drop") %>%
  mutate(BIOME = str_replace_all(
    T_MAPCODE,c("AT.*" = "Albany Thicket",
                "AZ.*" = "Azonal",
                "CB.*" = "IOCB", 
                "D.*"  = "Desert",
                "FF.*" = "Fynbos", 
                "FR.*" = "Fynbos",
                "FS.*" = "Fynbos",
                "FO.*" = "Forest", 
                "G.*" = "Grassland",
                "NK.*" = "Nama-Karoo", 
                "NOTVEG.*" = "NOTVEG",
                "SK.*" = "Succulent Karoo",
                "SV.*" = "Savanna"))) %>%
  mutate(LC7_text = case_when(
    LC == 1 ~ "Natural", 
    LC == 2 ~ "Secondary-natural", 
    LC == 3 ~ "Artificial-waterbody", 
    LC == 4 ~ "Built-up", 
    LC == 5 ~ "Cropland", 
    LC == 6 ~ "Mine",
    LC == 7 ~ "Plantation")) %>%
  relocate(BIOME, .after = T_MAPCODE)%>%
  mutate(km2 = n*400/1000000) #for test use 160000 for real use 400

write.csv(lc7_rall, file = "outputs/lc7_rall.csv") # The csv is used in other workflows (consider using .rds format for within R sharing as it maintains tibble formatting) 
```

### Summarise to test

```{r}
# Summarise to test
#lc7_rall <- read_csv("outputs/lc7_rall.csv", show_col_types = FALSE)
lc7_biome <- lc7_rall %>%   
  group_by(YEAR, LC7_text) %>%
  summarise(tot_km2 = sum(km2),.groups = 'drop') %>%
  pivot_wider(names_from = YEAR, values_from = tot_km2)


```
