---
bibliography: references.bib
---

### Load required packages

```{r}
#load packages   
library(terra)  
library(tidyverse) 
library(sf) 
library(PROJ) 
```

### Data sources & import:

1.  Little Karoo degradation map developed in 2009 [@thompson2009]. The severely degraded class from this data is estimated to be equivalent to 80 severity on RLE Criterion D - factors include overgrazing / browsing resulting in severe reduction in shrub canopy cover, changes in species composition (loss of palatable species usually dominant), increase in bare ground fraction (with soil loss). Most impacts occurred more than 50 years ago, and in this arid region the shrubland does not recover naturally over time - rather it enters an alternative stable state - bare ground with annual grass and herbs following rainfall events - limited perennial cover).

2.  Eastern Little Karoo degradation map (single ecosystem type covered) prepared by Jan and Annelize Vlok based on field surveys and aerial photographs (Vlok and Vlok pers com, 2020). Severely overgrazed / browsed areas (severity of \> 80%), with evidence (aerial photos) that the degradation occured in the last 50years).

3.  National Land Cover (7 class version prepared by SANBI) 2022.

4.  National Vegetation Map 2024 version, vector data (ESRI file geodatabase) January 2025 curated by SANBI [@rutherford2006; @Dayaram2019].

```{r}
# Add LK Data (prepared in ARCGIS PRO), where 0 = inside AOI but no value assigned, 8 = severely degraded 80%). 
LK <- rast("C:/Users/skownoa/Dropbox/NBAwork/RLEworking/Alternative_LC_DEG/littlekaroo_deg2.tif")
ELK <- rast("C:/Users/skownoa/Dropbox/NBAwork/RLEworking/Alternative_LC_DEG/easternlittlekaroo_deg2.tif")
# Add national LC, resample to match 20m AEA grid used for supplementary analysis
nlc <- rast("C:/Users/skownoa/Dropbox/NBAwork/NLC_change/nlc2022_7class.tif" )
nlc <- resample(nlc, LK, method = "near") #  match extent 

# Add vegetation data  
veg <- vect("C:/GIS/NBA/VegMap/VEGMAP2024/NVM2024final/NVM2024final.gdb", "NVM2024Final_IEM5_12_07012025")  

# Clean up non veg names 
veg$T_MAPCODE <- ifelse(veg$T_Name == "Non-terrestrial (estuarine habitat)" | veg$T_Name == "Non-terrestrial (Aquatic)", "NOTVEG", veg$T_MAPCODE)   
# rasterise the veg vector 
veg_r <- rasterize(veg, nlc, "T_MAPCODE")  
```

## Analysis

*Cross-tabulate land cover time series with vegetation.*

Step 1. Stack the land cover for each time period with the vegetation data.

Step 2. Cross-tabulate each raster stack and add a column denoting the YEAR (from Land Cover) to the output table (in tibble format). Rename each output column as "LC" to allow binding later.

```{r}
# Analysis #Step 1-2.   # get LC  at 1990 time period 
lk_rall  <- as_tibble(crosstab(c(nlc, veg_r, LK, ELK)))   
write.csv(lk_rall, file = "outputs/lk_rall.csv")

# Import cross tab output to avoid running spatial analysis 
#lk_rall <- read.csv("C:/Rdata/LCC_terr/outputs/lk_rall.csv")

# Calculate extent of each vegtype within the specific AOI 
lk_list <- lk_rall %>%
  filter(littlekaroo_deg2 != 99, T_MAPCODE != "NOTVEG") %>%
  group_by(T_MAPCODE) %>%
  summarise(ext_lk = sum(n)*400/1000000)

# clean up names and add km2
lk_rall  <- lk_rall  %>%    
  mutate(km2 = n*400/1000000) %>% # pixel size is 20m
  rename(LC = nlc2022_7class) %>%
  rename(LK = littlekaroo_deg2) %>%
  rename(ELK = easternlittlekaroo_deg2)

lk_sum <- lk_rall %>%
  filter(T_MAPCODE != "NOTVEG") %>%
  mutate(LCcomb = case_when(
                            ELK == 8 & LC == 1 ~ "deg80", 
                            ELK == 0 & LC == 1 ~ "nat", 
                            ELK == 0 & LC > 1 ~ "notnat", 
                            LK == 8 & LC == 1 ~ "deg80",
                            LK == 0 & LC == 1 ~ "nat",
                            LK == 0 & LC > 1 ~ "notnat",
                            LC == 1 ~ "nat", 
                            LC > 1 ~ "notnat", 
                            TRUE ~ "unknown")) %>%
  group_by(T_MAPCODE, LCcomb) %>%
  summarise(km2tot = sum(km2)) %>%
  pivot_wider(names_from = LCcomb, values_from = km2tot) %>%
    mutate(prpdeg = deg80/(nat+deg80)) %>% # cond of remnants (D1)
    mutate(prplost = notnat/(nat+notnat+deg80)) %>% # ext of remnants (A3)
    mutate(D3 = case_when(prpdeg >= 0.9 ~ "CR",
                   prpdeg >= 0.7 ~ "EN",
                   prpdeg >= 0.5 ~ "VU",
                  TRUE ~ "LC")) %>%
    left_join(lk_list, by = "T_MAPCODE") %>%
   mutate(ext_lk = ext_lk/(nat + notnat + deg80))

write.csv(lk_sum, file = "outputs/lk_sum.csv")

elk_sum <- lk_rall %>%
filter(T_MAPCODE != "NOTVEG", T_MAPCODE == "SKv11", n != 0)  %>% 
         mutate(LCcomb = case_when(
                            ELK == 8 | LK == 8  ~ "deg80",
                            ELK == 0 & LC == 1 ~ "nat", 
                            ELK == 0 & LC > 1 ~ "notnat", 
                            TRUE ~ "unknown")) %>%
  group_by(T_MAPCODE, LCcomb) %>%
  summarise(km2tot = sum(km2)) %>%
  pivot_wider(names_from = LCcomb, values_from = km2tot) %>%
    mutate(prpdeg = deg80/(nat+deg80)) %>% # cond of remnants (D1)
    mutate(prplost = notnat/(nat+notnat+deg80)) %>% # ext of remnants (A3)
    mutate(D1 = case_when(prpdeg >= 0.8 ~ "CR",
                   prpdeg >= 0.5 ~ "EN",
                   prpdeg >= 0.3 ~ "VU",
                  TRUE ~ "LC")) 

write.csv(elk_sum, file = "outputs/elk_sum.csv")
  
```
